---
title: "Variations on a `...` theme"
subtitle: "Proposals for `...` in R"
author: "Doug Kelkhoff @ DSC 2019"
date: "2019-09-18"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r, echo = FALSE}
suppressPackageStartupMessages(library(dplyr))
```





## First, an ode to `...`

```{r}
dots_how_do_i_love_thee <- function(...) {
  cat("let me count the ways... \n")
  cat(paste(..., sep = "\n"))
}
```

```{r}
dots_how_do_i_love_thee("really flexible", "mirrors natural language", "intuitive")
```





## The `...` are really handy, but can get unwieldy

- Often need to be collapsed into list, thwarting laziness
- Comes with a cohort of unintuitive operators/functions
- No way to operate on named ... elements without evaluation (?)
- Paradigm is only applicable within function calls

```{r}
dots_how_do_i_love_thee <- function(...) {
  cat("let me count the ways... \n")
  cat(paste0("  ", seq(1, ...length()), ". ", list(...), collapse = "\n"))
}

dots_how_do_i_love_thee("really flexible", "mirrors natural language", "intuitive")
```

### Some `...` syntax

`...`, `..n`, `...length()`, `...elt()`






## Variadic functions popular in tons of languages

There are plenty of implementations of this feature, some with quite interesting
edge case handling

[Rosetta Code: Variatic Functions](www.rosettacode.org/wiki/Variadic_function)

### `python`

```{r, echo = FALSE, results = 'asis'}
cat("<pre class='prettyprint lang-python'>

def my_func(*args, **kwargs): 
    other_func(*args, **kwargs)
    
</pre>")
```

### `julia`

```{r, echo = FALSE, results = 'asis'}
cat("<pre class='prettyprint lang-julia'>

function my_func(args...; kwargs...)
    other_func(args...; kwargs...)
end
 
</pre>")
```






## Ellipsis leverage language familiarity

### _University of Oxford Style Guide_: 

- "The quick brown fox jumps over the lazy dog… And if they have not died, they are still alive today."
- "It is not cold… it is freezing cold."

[Wikipedia "Ellipsis"](https://en.wikipedia.org/wiki/Ellipsis#In_British_English)

### They map well to use in verbal languages

- you might expect a list to continue...
- if followed by an ellipsis
- ...or a long pause before an end

### To generalize...

- `something ...`: more to come
- `... something`: finishing something







# How far can we push this familiarity?

## Prompt

<font size="6" style = "line-height: 1.5">
_Are there ways we can extend our intuition for `...` to other elements of the
language?_
</font>





## **Idea 1**: Ellipsis unpacking

### Composing function calls in R is high bar for new users

```{r, eval = FALSE}
args <- list("gone!", sep = ", ")
cat(do.call(paste, append(list("going", "going"), args)))
```

- `do.call` assumes pretty strong familiarity of first class functions
- Argument lists must be composed dynamically

### Instead, could arguments lists be unpacked directly into a call

```{r, eval = FALSE}
args <- list("gone!", sep = ", ")
cat(paste("going", "going", ...args))
```

- Retains familiar function call structure
- Syntactically cleaner
- Extends `...` paradigm






## **Idea 2**: Named Ellipsis Parameters

Taking a page from **Julia**, allow naming of a "rest" argument

```{r, eval = FALSE}
example <- function(dots...) {
  # allow for easier subsetting, manipulation without 
  # collapsing to list(...) or handling eval in parent frame
  cat(...dots, sep = ", ")
}
```

`function(rest...) class(rest)  # possibly a list of unevaluated promises?`






## **Idea 2**: Named Ellipsis Parameters

Taking a page from **Julia**, allow naming of a "rest" argument

```{r, eval = FALSE}
example <- function(dots...) {
  # allow for easier subsetting, manipulation without 
  # collapsing to list(...) or handling eval in parent frame
  
  dots <- dots[!names(dots) %in% "sep"]
  cat(...dots, sep = ", ")
}
```

But we still need to handle repeated argument names to avoid `...` induced errors






## **Idea 3**: Better yet, allow repeated arguments

Use ellipses position to indicate precedence

If an argument is passed in ellipses (not explicitly named twice), allow the
most recent argument to take precedence.

```{r, eval = FALSE}
example <- function(...) {
  # fix the 'sep' field regardless of what's in dots
  cat(..., sep = ", ")
  #  set a default that is overwritten if present in dots
  cat(sep = ", ", ...)
}
```

_`julia` implements ellipsis passing as a special case where rightmost argument is used_






## **Idea 4**: Parital Function Application

Appending ellipsis after a function to indicate that it should return a
partially applied function instead of the call result

```{r, eval = FALSE}
newline_cat <- cat(sep = "\n")...
newline_cat("word", "per", "line")
```

```{r, echo = FALSE}
cat("word", "per", "line", sep = "\n")
```

- Retains formals
- Could propegate documentation
- Especially helpful for tab completions






## _Aside_: A mental model for argument unpacking

```
my_function <- function(a, b, c, d, e = 4, dots...) <stuff> 
args <- list(1, b = 2, c = 3)
my_function(0, a = 2, ...args)
```

#### Step 1: Consider the function formals

```{r, echo = FALSE, results = 'asis'}
cat("<pre class='prettyprint lang-r'>")
cat("# <----- what I passed ----->       <--- my_function formals -->", "\n")
cat("  (0, a = 2, 1, b = 2, c = 3)  ==>  (a, b, c, d, e = 4, dots...)")
cat("</pre>")
```

#### Step 2: Fill in formal default values

```{r, echo = FALSE, results = 'asis'}
cat("<pre class='prettyprint lang-r'>")
cat("(0, a = 2, 1, b = 2, c = 3, <strong>e = 4</strong>)")
cat("</pre>")
```

#### Step 3: Align named arguments

```{r, echo = FALSE, results = 'asis'}
cat("<pre class='prettyprint lang-r'>")
cat("(<strong>a = 2, b = 2, c = 3, e = 4</strong>, 0, 1)")
cat("</pre>")
```

#### Step 4: Backfill positional arguments

```{r, echo = FALSE, results = 'asis'}
cat("<pre class='prettyprint lang-r'>")
cat("(a = 2, b = 2, c = 3, e = 4, <strong>d = 0, dots... = 1</strong>)")
cat("</pre>")
```






## **Idea 5**: Return list unpacking

Mirror list unpacking into function calls with unpacking into assigned return
values

Syntactically parallels function parameter aliasing

```{r, eval = FALSE}
(x, y, z...) <- list(w = 1, x = 2, y = 3, z = 3)
```

```
> x
## [1] 2

> y
## [1] 3

> z
## $w
## [1] 1
## 
## $z
## [1] 3
```


## **Idea 5**: Return list unpacking... considerations

- Can we get rid of the `()`'s?

```{r, eval = FALSE}
x, y, z... <- list(w = 1, x = 2, y = 3, z = 3)
```

- Requiring unpacking syntax?

```{r, echo = FALSE, eval = FALSE}
(x, y, z...) <- ...list(w = 1, x = 2, y = 3, z = 3)
```

```{r, echo = FALSE, results = "asis"}
# r code for rendering text with emphasis for slides

cat("<pre class='prettyprint lang-r'>")
cat("(x, y, z...) <- <strong>...</strong>list(w = 1, x = 2, y = 3, z = 3)")
cat("</pre>")
```

- Allowing mapping list names to target object names?

```{r, echo = FALSE, eval = FALSE}
(a = x, b = y, c...) <- list(w = 1, x = 2, y = 3, z = 3)
```

```{r, echo = FALSE, results = "asis"}
# r code for rendering text with emphasis for slides

cat("<pre class='prettyprint lang-r'>")
cat("(<strong>a = x, b = y</strong>, c...) <- list(w = 1, x = 2, y = 3, z = 3)")
cat("</pre>")
```

- Should the `rest...` contain _just_ the remaining values or the entire list? 
- Should it be possible to get both?
- Worthwhile having a thunk syntax (`_`) to throw away list elements? 

```{r, echo = FALSE, results='asis'}
cat("<pre class='prettyprint lang-haskell'>")
cat("
# taking a page from Haskell
# getting both entirety of list (list) and sub-components head (x) & remainder (xs)
f list@(x:xs) = ...
")
cat("</pre>")
```






## **Idea 6:** Anonymous function shorthand

Draw inspiration from the `purrr` package to create an unambiguous lambda
function syntax

```{r, eval = FALSE}
# function(...) ..1 + ..2
~> ..1 + ..2
```

```{r, eval = FALSE}
# function(x, y, ...) x + y
x, y ~> x + y
```

- reminescent of `purrr`-style lambda function syntax
- disambiguates lambdas from formulas (`:symbol` shorthand for `name`?)

### Another alternative for "partial application" 

```{r, eval = FALSE}
new_cat <- ~>cat(sep = ", ", ...)
```

_retaining formals and docs require special handling for singular call_






# Why are these conveniences important to the longevity of the language?

## Enter the `Tidyverse`

The tidyverse, and its incredible mindshare, has begun to implement many of
these conveniences. 

New users have trouble tracking tidyverse-specific syntax

### Argument unpacking (and unquoting) `!!!`

```{r, eval = FALSE}
my_mutations <- list(new_var = "new_var")
mtcars %>% mutate(!!!my_mutations)
```

### `purrr`-style lambdas (now in `rlang`) `~`

```{r, eval = FALSE}
mtcars %>% mutate_at(vars(cyl), ~ . * 2)
```

### `ggplot2` symbol representation

```{r, eval = FALSE}
ggplot(mtcars) + 
  aes_(~mpg, ~wt + wt) +  # requires parsing of ~rhs
  geom_point()
```






## Reconciling the Tidyverse

Some of the proposed syntax can be used to bring consistency to the
tidyverse/base bifurcation

### Argument unpacking

```{r, eval = FALSE}
my_mutations <- list(new_var = "new_var")
mtcars %>% mutate(...my_mutations)
```

_handles unpacking, but not unquoting_

### Lambdas

```{r, eval = FALSE}
mtcars %>% mutate_at(vars(cyl), x ~> x * 2)
```

### Name Notation

```{r, eval = FALSE}
ggplot(mtcars) + 
  aes(:mpg, :wt + :wt) + 
  geom_point()
```






## Closing Thoughts

```{r, eval = FALSE}
dots_how_do_i_love_thee <- function(dots...) {
  class(dots)                             # - list of unevaluateded promise?
  names(dots)                             # - operate on list without evaluating
  (pdots, ndots) <- ...split_named(dots)  # - define helpers that keep laxiness
  cat("let me count the ways... \n")
  cat(paste0("  ", seq_along(pdots), ". ", pdots, collapse = "\n", ...ndots, sep = " "))
}
```

- `...` is awesome syntactic feature in R, balancing usability against readability
- Offers opportunities for expanding on paradigm
- Developers benefit from handling `rest...` args without breaking laziness
- Users benefit from consistency of `...` arguments
- Consistency among package implementations reduces bucketing of expectations (e.g. `tidyverse` vs `base`)






## Questions & Discussion

### Special Thanks

Michael Lawrence, Gabe Becker

Genentech, Roche
